---
title: "Introduction to Shiny: Building a Birthday App"
author: "Dr Melanie Smith"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

# Introduction to Shiny

## What is Shiny?

Shiny is an R package that makes it easy to build interactive web applications directly from R. Despite being a powerful framework, Shiny is accessible to R users without web development experience. In this tutorial, we'll build a simple application that:

1. Accepts a JPEG image upload
2. Lets users select their birth date using a calendar widget
3. Displays the day of the week they were born on

## Prerequisites

Before we begin, make sure you have the necessary packages installed:

```{r install-packages}
# Install required packages if you don't have them
if (!require("shiny")) install.packages("shiny")
if (!require("lubridate")) install.packages("lubridate")
```

# Understanding the Structure of a Shiny App

Every Shiny app consists of two main components:

1. **UI (User Interface)**: Controls the layout and appearance of the app
2. **Server**: Contains the logic that tells the app how to process inputs and create outputs

A basic Shiny app structure looks like this:

```{r basic-structure}
library(shiny)

# Define UI
ui <- fluidPage(
  # UI elements go here
)

# Define server logic
server <- function(input, output) {
  # Server logic goes here
}

# Run the application 
shinyApp(ui = ui, server = server)
```

# Building Our Birthday App Step by Step

Now, let's build our birthday app piece by piece:

## Step 1: Creating the Basic Framework

First, let's set up the basic structure of our app:

```{r framework}
library(shiny)
library(lubridate)  # For working with dates

# Define UI
ui <- fluidPage(
  titlePanel("Birthday Day Calculator"),
  
  sidebarLayout(
    sidebarPanel(
      # Our input controls will go here
    ),
    
    mainPanel(
      # Our outputs will go here
    )
  )
)

# Define server logic
server <- function(input, output) {
  # Our server logic will go here
}

# Run the application 
shinyApp(ui = ui, server = server)
```

In this framework:

- `titlePanel()` creates a header for our app  
- `sidebarLayout()` divides our app into two sections: a sidebar for inputs and a main panel for outputs  
- `sidebarPanel()` is where we'll put our controls (image upload and date selector)  
- `mainPanel()` is where we'll display our results (image and text)  

## Step 2: Adding Input Controls

Now, let's add the input controls to our app:

```{r input-controls}
ui <- fluidPage(
  titlePanel("Birthday Day Calculator"),
  
  sidebarLayout(
    sidebarPanel(
      # Upload JPEG image
      fileInput("imageFile", "Upload your JPEG image:",
                accept = c("image/jpeg", "image/jpg")),
      
      # Date input for selecting birth date
      dateInput("birthDate", "Select your birth date:", 
                value = Sys.Date(),
                format = "yyyy-mm-dd"),
      
      # Submit button
      actionButton("submitBtn", "Submit")
    ),
    
    mainPanel(
      # Our outputs will go here
    )
  )
)
```

Let's break down the input controls:

1. `fileInput("imageFile", "Upload your JPEG image:")`: Creates a file upload control
   - The first argument `"imageFile"` is the input ID (we'll use this to access the file in the server)
   - The second argument is the label displayed to users
   - `accept = c("image/jpeg", "image/jpg")` limits uploads to JPEG files only

2. `dateInput("birthDate", "Select your birth date:")`: Creates a calendar widget
   - `"birthDate"` is the input ID 
   - `value = Sys.Date()` sets the default value to today's date
   - `format = "yyyy-mm-dd"` specifies how the date should be displayed

3. `actionButton("submitBtn", "Submit")`: Creates a button that users will click to submit their information

## Step 3: Adding Output Elements

Now let's add the elements that will display our outputs:

```{r output-elements}
ui <- fluidPage(
  titlePanel("Birthday Day Calculator"),
  
  sidebarLayout(
    sidebarPanel(
      # Upload JPEG image
      fileInput("imageFile", "Upload your JPEG image:",
                accept = c("image/jpeg", "image/jpg")),
      
      # Date input for selecting birth date
      dateInput("birthDate", "Select your birth date:", 
                value = Sys.Date(),
                format = "yyyy-mm-dd"),
      
      # Submit button
      actionButton("submitBtn", "Submit")
    ),
    
    mainPanel(
      # Display the uploaded image
      imageOutput("imageDisplay"),
      
      # Display the result text
      h3(textOutput("resultText")),
      
      # Optional: Display the selected date
      verbatimTextOutput("dateInfo")
    )
  )
)
```

The output elements are:

1. `imageOutput("imageDisplay")`: Creates a container for our uploaded image
   - `"imageDisplay"` is the output ID that we'll use in the server logic

2. `h3(textOutput("resultText"))`: Creates a heading (h3) containing our result text
   - `textOutput("resultText")` will display our "You were born on a ___" message
   - Wrapping it in `h3()` makes it display as a larger heading

3. `verbatimTextOutput("dateInfo")`: Creates a verbatim text output for the selected date
   - This will display the selected date in a fixed-width font

## Step 4: Adding Server Logic

Now, let's write the server logic that processes inputs and creates outputs:

```{r server-logic}
server <- function(input, output) {
  
  # Display the uploaded image
  output$imageDisplay <- renderImage({
    req(input$imageFile)
    
    # Validate that the file is a jpeg
    if (!grepl("jpeg|jpg", input$imageFile$type, ignore.case = TRUE)) {
      return(NULL)
    }
    
    # Return a list with the image info
    list(
      src = input$imageFile$datapath,
      contentType = input$imageFile$type,
      width = "100%",
      alt = "Uploaded image"
    )
  }, deleteFile = FALSE)
  
  # Get the day of week
  dayOfWeek <- eventReactive(input$submitBtn, {
    req(input$birthDate)
    weekdays(input$birthDate)
  })
  
  # Create the output text
  output$resultText <- renderText({
    req(dayOfWeek())
    paste("You were born on a", dayOfWeek())
  })
  
  # Display the selected date (optional)
  output$dateInfo <- renderText({
    req(input$birthDate)
    paste("Selected date:", format(input$birthDate, "%B %d, %Y"))
  })
}
```

Let's break down the server logic:

1. **Image Display Logic**:
   ```r
   output$imageDisplay <- renderImage({
     req(input$imageFile)
     
     # Validate that the file is a jpeg
     if (!grepl("jpeg|jpg", input$imageFile$type, ignore.case = TRUE)) {
       return(NULL)
     }
     
     # Return a list with the image info
     list(
       src = input$imageFile$datapath,
       contentType = input$imageFile$type,
       width = "100%",
       alt = "Uploaded image"
     )
   }, deleteFile = FALSE)
   ```
   - `req(input$imageFile)` ensures the logic only runs when a file is uploaded
   - The validation checks if the file has a JPEG extension
   - `renderImage()` expects a list with specific properties (src, contentType, etc.)
   - `deleteFile = FALSE` prevents Shiny from deleting the file after display

2. **Day of Week Calculation**:
   ```r
   dayOfWeek <- eventReactive(input$submitBtn, {
     req(input$birthDate)
     weekdays(input$birthDate)
   })
   ```
   - `eventReactive()` makes the calculation only happen when the submit button is clicked
   - `weekdays()` is a function from base R that extracts the day of the week from a date

3. **Result Text Creation**:
   ```r
   output$resultText <- renderText({
     req(dayOfWeek())
     paste("You were born on a", dayOfWeek())
   })
   ```
   - `renderText()` creates text output
   - `req(dayOfWeek())` ensures the calculation has been performed
   - `paste()` combines the text with our calculated day

4. **Date Info Display**:
   ```r
   output$dateInfo <- renderText({
     req(input$birthDate)
     paste("Selected date:", format(input$birthDate, "%B %d, %Y"))
   })
   ```
   - This shows the selected date in a more readable format

## Step 5: Putting It All Together

Now, let's put everything together to create our complete Shiny app:

```{r complete-app}
library(shiny)
library(lubridate)

# Define UI
ui <- fluidPage(
  titlePanel("Birthday Day Calculator"),
  
  sidebarLayout(
    sidebarPanel(
      # Upload JPEG image
      fileInput("imageFile", "Upload your JPEG image:",
                accept = c("image/jpeg", "image/jpg")),
      
      # Date input for selecting birth date
      dateInput("birthDate", "Select your birth date:", 
                value = Sys.Date(),
                format = "yyyy-mm-dd"),
      
      # Submit button
      actionButton("submitBtn", "Submit")
    ),
    
    mainPanel(
      # Display the uploaded image
      imageOutput("imageDisplay"),
      
      # Display the result text
      h3(textOutput("resultText")),
      
      # Optional: Display the selected date
      verbatimTextOutput("dateInfo")
    )
  )
)

# Define server logic
server <- function(input, output) {
  
  # Display the uploaded image
  output$imageDisplay <- renderImage({
    req(input$imageFile)
    
    # Validate that the file is a jpeg
    if (!grepl("jpeg|jpg", input$imageFile$type, ignore.case = TRUE)) {
      return(NULL)
    }
    
    # Return a list with the image info
    list(
      src = input$imageFile$datapath,
      contentType = input$imageFile$type,
      width = "100%",
      alt = "Uploaded image"
    )
  }, deleteFile = FALSE)
  
  # Get the day of week
  dayOfWeek <- eventReactive(input$submitBtn, {
    req(input$birthDate)
    weekdays(input$birthDate)
  })
  
  # Create the output text
  output$resultText <- renderText({
    req(dayOfWeek())
    paste("You were born on a", dayOfWeek())
  })
  
  # Display the selected date (optional)
  output$dateInfo <- renderText({
    req(input$birthDate)
    paste("Selected date:", format(input$birthDate, "%B %d, %Y"))
  })
}

# Run the application 
shinyApp(ui = ui, server = server)
```

# How to Run the App

To run the app:

1. Create a new R script or RMarkdown file
2. Copy the complete code from above
3. Save the file with a name like "birthday_app.R" or "app.R"
4. Click the "Run App" button in RStudio (or use `shiny::runApp()` in the console)

# Key Shiny Concepts to Understand

## Reactivity

Reactivity is a key concept in Shiny. When an input value changes, any outputs that depend on it automatically update. There are several important reactive functions:

- `reactive()`: Creates a reactive expression that updates when its dependencies change
- `eventReactive()`: Creates a reactive expression that updates only when a specific event occurs
- `observeEvent()`: Performs an action when a specific event occurs
- `req()`: Ensures that a value is available before proceeding

In our app, we used `eventReactive()` to calculate the day of the week only when the submit button is clicked, rather than automatically when the date changes.

## Input and Output IDs

Notice how we use IDs to connect UI elements to server logic:

1. In the UI, we create input controls with IDs like `"imageFile"` and `"birthDate"`
2. In the server, we access these inputs using `input$imageFile` and `input$birthDate`
3. Similarly, we define output IDs in the UI (`"imageDisplay"`, `"resultText"`) and set these outputs in the server using `output$imageDisplay` and `output$resultText`

This ID system is how Shiny knows which elements are connected.

## Render Functions

Different types of outputs require different render functions:

- `renderText()`: For text output
- `renderImage()`: For images
- `renderPlot()`: For plots
- `renderTable()`: For tables
- `renderUI()`: For dynamic UI elements

Each render function pairs with a corresponding output function in the UI.

# Challenge Exercises

Now that you understand the basics, try these exercises to extend the app:

1. Add error handling for when no image is uploaded
2. Modify the app to display the birth month as well as the day of the week
3. Add a feature that calculates the user's age in years, months, and days
4. Change the app's appearance using a theme from the `bslib` package

# Conclusion

In this tutorial, you've learned:

1. The basic structure of a Shiny app
2. How to create input controls and output elements
3. How to write server logic to process inputs and create outputs
4. How reactivity works in Shiny

Shiny is a powerful tool for creating interactive web applications directly from R. With the knowledge you've gained from this tutorial, you can start building your own interactive applications for data analysis, visualization, and more.

# Resources for Further Learning

- [Shiny Official Documentation](https://shiny.rstudio.com/)
- [Shiny Gallery](https://shiny.rstudio.com/gallery/)
- [Mastering Shiny](https://mastering-shiny.org/) by Hadley Wickham
- [RStudio Shiny Cheatsheet](https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)