---
title: "Installation and Setup"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      eval = FALSE)
```


Follow the install instructions below to prepare for the workshop.

## R and RStudio

Ensure that RStudio is installed. For this workshop we recommend running R >= 4.1.

To check the version of R you are using, start RStudio and the first thing that appears on the terminal indicates the version of R you are running. Alternatively, you can type `sessionInfo()` in the console, which will also display which version of R you are running.

## XQuartz - Recommended for macOS users

For those with MacOS it is recommended that you have XQuartz installed as it is required by some of the packages we will be using. 

To check if this is already installed on your machine go to launchpad and search for XQuartz. Alternatively, open your applications directory and perform a search (XQuartz is often installed within the `utilities` directory of your `applications` folder).

If XQuartz is not installed, you'll need to download and install it from the [XQuartz website](https://www.xquartz.org/).

## R packages

Before we begin, we need to have the shiny package installed, as well as
a few other key packages we will be using. Run the following commands in your R console.

```{r set-up, eval=FALSE}
# Install required packages if you don't have them
if (!require("shiny")) install.packages("shiny") 
if (!require("lubridate")) install.packages("lubridate") 
if (!require("tidyverse")) install.packages("tidyverse") 
if (!require("here")) install.packages("here") 
if (!require("DT")) install.packages("DT") 
if (!require("pheatmap")) install.packages("pheatmap") 
if (!require("plotly")) install.packages("plotly") 
if (!require("heatmaply")) install.packages("heatmaply") 
```

To ensure you successfully installed Shiny, try running one of the demo apps.

```{r set-up check, eval=FALSE}
library(shiny)
runExample("01_hello")
```

If the example app is running, press Escape to close the app, and you are ready to build your first Shiny app!

## Data and Examples

During the workshop we will be building several shiny apps to demonstrate key concepts and features for R Shiny. To help with this we have provided some pre-compiled data that will be used, along with some example apps and R markdown files with the tutorial instructions we will be following.

Download the workshop material [here](https://drive.google.com/file/d/1SAo9DZxR_GR1mz7VJXU9D-sp3KSaLHHq/view?usp=sharing) by clicking on the download button (top right). This will download a zip directory of all materials.

- Move the download to a desired location on your computure and unzip the file.  
- Click on the `shiny_workshop.Rproj` file to open up the R project in RStudio.  
- Navigate to the tutorials folder, and click the `02_birthday_app.Rmd` file to open up the tutorial in RStudio.  
- You can also, navigate to the docs folder, and click on the `02_birthday_app.html` file to open up the tutorial documentation in a web browser.  
